Airspace configuration decision based on Monte-Carlo tree search

### 2018-04-21 article/article.tex
(~) Simplified a lot layout
latex defaults are not that bad
(~) Corrections in intro

\documentclass[twoside,twocolumn]{article}

@@

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption}
\usepackage{booktabs}

\usepackage{lettrine}

\usepackage{enumitem}
\setlist[itemize]{noitemsep}

\usepackage{abstract}
% Set the "Abstract" text to bold
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
% Set the abstract itself to small italic text
\renewcommand{\abstracttextfont}{\normalfont\small\itshape}

% Allows customization of titles
\usepackage{titlesec}

% Headers and footers
\usepackage{fancyhdr}
% All pages have headers and footers
\pagestyle{fancy}
% Blank out the default header
% \fancyhead{}
% Blank out the default footer
\fancyfoot{}
% Custom header text
% \fancyhead[C]{Running title $\bullet$ May 2016 $\bullet$ Vol. XXI, No. 1}
% Custom footer text
\fancyfoot[RO,LE]{\thepage}
\usepackage{titling} % Customizing the title section

\usepackage{hyperref} % For hyperlinks in the PDF

\title{Airspace configuration decision based on Monte-Carlo tree search}
\author{Gabriel~Hondet, Beno\^{\i}t Viry}
\date{\today}

@@

% Move the title up
\setlength{\droptitle}{-4\baselineskip}
% Article title formatting
\pretitle{\begin{center}\Huge\bfseries}
% Article title closing formatting
  \posttitle{\end{center}}

\renewcommand{\maketitlehookd}{%
  \begin{abstract}

  \end{abstract}
}

%-----------------------------------------------------------------------------

\begin{document}

@@

\section*{Notations}
\begin{itemize}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(\mathcal{S}\): the set of available states;
  \item \(\mathcal{A}\): the set of all possible actions;
  \item \(\mathcal{A}(s)\): the available actions from a state \(s\);
  \item $s$: state;
  \item \(f \colon \mathcal{S} \times \mathcal{A} \to \mathcal{S}\):
    transition function giving the resulting states of action \(a\) executed
    from a state \(s\).
  \item $h \colon \mathcal{S} \to \mathbb{R}^+$: heuristic estimating the cost
    from a state \(s \in \mathcal{S}\) to a final state;
  \item \(\mathcal{N}\): the set of nodes
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item \(Q \colon \mathcal{N} \to \mathbb{R}^+\): the reward expectancy of a
    node;
  \item \(N \colon \mathcal{N} \to \mathbb{N}\): number of simulations running
    through or from a node;
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{itemize}

@@

Currently, configuration is mainly decided on the fly by chief officer (TODO).
This decision is based on the actual workload on each position. This intuitive
approach does not take into account future traffic. This papers aims at
providing a method offering a smoother workload during the day.

@@

As seen in~\cite{treimuth2016branch}
or~\cite{sergeeva2017dynamic}, a temporal sequence of configurations will be
provided. Two costs will be considered to create the sequence, namely the cost
associated to each configuration and transition. The former is based on the
workload estimated for one sector given a neural network
(see~\cite{gianazza2010forecasting}).

The sequence of configurations will be represented as a tree. The resulting
sequence will therefore be a path from the root to a leaf of this tree. This
problem is highly combinatorial (partitioning of the airspace). Knowing that a
sochastic tree search algorithm has been honing from a good
player~\cite{gelly2012grand} to the best one with AlphaGo on one of the most
highly combinatorial game wich is the game of Go entices us to use one for this
problem. The Monte Carlo tree search~\cite{browne2012survey} algorithm will be
used to fullfill this task.

\section{Previous related works}
\label{sec:previous_related_works}

@@

\subsubsection{Extending the search tree}
\paragraph{Selection (algorithm~\ref{alg:tree_pol})}
The aim of the selection is to choose from a set of actions, the most promising
one\footnote{i.e.\ possibly leading to the best evaluation} according to a
\emph{tree policy}. This type of problem is addressed by bandits methods where
a multi armed bandit aims to always activate the slot machine giving the highest
reward.

The bandit problem is applied to MCTS via the Upper Confidence Tree algorithm.
It uses the UCB (Upper Confidence Bound) as tree policy (introduced
in~\cite{kocsis2006bandit}). Let \(v\) be the node from which an action must be
selected to go deeper in the tree, \(a\) an action available from \(v\),
\(\mu_{v,a}\) the expected outcome from the node obtained by applying action
\(a\) on \(v\), \(T_v\) the number of simulations carried out from node \(v\)
(which includes any simulation from nodes in any subtree of \(v\)) and
\(T_{v,a}\) the number of simulations carried out from the node resulting from
the action \(a\) applied to node \(v\), \(\beta\) a constant. The selected node
is thus the one maximising the \emph{UCB1} equation
\begin{equation}
  \mu_{v,a} + 2 \beta \sqrt{\frac{2 \log T_v}{T_{v,a}}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficience in one player games or sequencing problems. An
alternative using the standard deviation of the outcome of the simulations is
proposed in~\cite{sebag2010fuse} called the \emph{UCB1-tuned} equation
\begin{equation}
  \mu_{v,a} +
  \sqrt{%
    \frac{\beta\log T_v}{t_{v,a}}
    \min\left( \frac{1}{4}, \sigma^2_{v,a} +
    \sqrt{%
      \frac{2 \log T_v}{T_{v,a}}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{$\pi$}
    \State{}\(v \gets \first(\pi)\)
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} | v' \text{\,children of\,} v \}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

>>>>

\documentclass[oneside,twocolumn]{article}

@@

%\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
%\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption}
%\usepackage{booktabs}
%
%\usepackage{lettrine}
%
%\usepackage{enumitem}
%\setlist[itemize]{noitemsep}
%
%\usepackage{abstract}
%% Set the "Abstract" text to bold
%\renewcommand{\abstractnamefont}{\normalfont\bfseries}
%% Set the abstract itself to small italic text
%\renewcommand{\abstracttextfont}{\normalfont\small\itshape}
%
%% Allows customization of titles
%\usepackage{titlesec}
%
%% Headers and footers
%\usepackage{fancyhdr}
%% All pages have headers and footers
%\pagestyle{fancy}
%% Blank out the default header
%% \fancyhead{}
%% Blank out the default footer
%\fancyfoot{}
%% Custom header text
%% \fancyhead[C]{Running title $\bullet$ May 2016 $\bullet$ Vol. XXI, No. 1}
%% Custom footer text
%\fancyfoot[RO,LE]{\thepage}
%\usepackage{titling} % Customizing the title section

\title{Airspace configuration decision based on Monte-Carlo tree search}
\author{Gabriel~Hondet, Beno\^{\i}t Viry}
\date{\today}
\pagestyle{headings}

@@

%% Move the title up
%\setlength{\droptitle}{-4\baselineskip}
%% Article title formatting
%\pretitle{\begin{center}\Huge\bfseries}
%% Article title closing formatting
%  \posttitle{\end{center}}

%\renewcommand{\maketitlehookd}{%
%  \begin{abstract}
%
%  \end{abstract}
%}

%-----------------------------------------------------------------------------
\begin{document}

@@

\begin{abstract}
  A nice abstract
\end{abstract}

\section*{Notations}
\begin{compactitem}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(\mathcal{S}\): the set of available states;
  \item \(\mathcal{A}\): the set of all possible actions;
  \item \(\mathcal{A}(s)\): the available actions from a state \(s\);
  \item $s$: state;
  \item \(f \colon \mathcal{S} \times \mathcal{A} \to \mathcal{S}\):
    transition function giving the resulting states of action \(a\) executed
    from a state \(s\).
  \item $h \colon \mathcal{S} \to \mathbb{R}^+$: heuristic estimating the cost
    from a state \(s \in \mathcal{S}\) to a final state;
  \item \(\mathcal{N}\): the set of nodes
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item \(Q \colon \mathcal{N} \to \mathbb{R}^+\): the reward expectancy of a
    node;
  \item \(N \colon \mathcal{N} \to \mathbb{N}\): number of simulations running
    through or from a node;
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{compactitem}

@@

Currently, configuration is mainly decided on the fly by chief officer (TODO).
This decision is based on the actual workload on each position. In this
approach, future workload estimation is based on the controller's feelings and
it therefore lacks a workload prediction tool.
This papers aims at providing this tool as well as a method offering a smoother
workload during the day.

@@

In this paper, a temporal sequence of configuration is provided, as done
in~\cite{treimuth2016branch} or~\cite{sergeeva2017dynamic}.
Two costs are considered to create the sequence, namely the cost
associated to each configuration and a transition cost. The former is based on
the workload estimated for one sector given a neural network
(see~\cite{gianazza2010forecasting}).

The sequence of configurations is represented as a tree. The resulting
sequence will therefore be a path from the root to a leaf of this tree. This
problem is highly combinatorial (partitioning of the airspace). Since stochastic
tree search algorithms, combined with deep neural networks have proved
themselves worthy by outranking the best human player of one of the most highly
combinatorial game which is the game of Go, the Monte Carlo tree
search~\cite{browne2012survey} algorithm is used in this paper to fullfill the
previously mentioned task.

\section{Previous related works}\label{sec:previous_related_works}

@@

\subsubsection{Extending the search tree}
\paragraph{Selection (algorithm~\ref{alg:tree_pol})}
The aim of the selection is to choose from a set of actions, the most promising
one\footnote{i.e.\ possibly leading to the best evaluation} according to a
\emph{tree policy}. This type of problem can be solved by bandits methods. Those
methods consist in, given a bandit in front of several slot machines (multi
armed bandit), decide which machine will bring the highest reward knowing the
past result. The objective of bandits methods is thus to maximise the reward and
minimise the regret of not playing the best machine.

The bandit problem has been applied to MCTS via the Upper Confidence Tree
algorithm in~\cite{kocsis2006bandit} using the UCB1 equation~\ref{eq:ucb1}.
Let \(v\) be the node from which an action must be
selected to go deeper in the tree, \(a\) an action available from \(v\),
\(\mu_{v,a}\) the expected outcome from the node obtained by applying action
\(a\) on \(v\), \(T_v\) the number of simulations carried out from node \(v\)
(which includes any simulation from nodes in any subtree of \(v\)) and
\(T_{v,a}\) the number of simulations carried out from the node resulting from
the action \(a\) applied to node \(v\), \(\beta\) a constant. The selected node
is thus the one maximising the \emph{UCB1} equation
\begin{equation}
  \label{eq:ucb1}
  \mu_{v,a} + 2 \beta \sqrt{\frac{2 \log T_v}{T_{v,a}}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficiency in one player games or sequencing problems. An
alternative using the standard deviation of the outcome of the simulations is
proposed in~\cite{sebag2010fuse} called the \emph{UCB1-tuned} equation
\begin{equation}
  \mu_{v,a} +
  \sqrt{%
    \frac{\beta\log T_v}{T_{v,a}}
    \min\left( \frac{1}{4}, \sigma^2_{v,a} +
    \sqrt{%
      \frac{2 \log T_v}{T_{v,a}}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{$\pi$}
    \State{}\(v \gets \first(\pi)\)
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} | v' \text{\,children of\,} v \}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
